### 2020-11-02 forms/de_AT/1.tex
Initial commit

\documentclass{article}
\usepackage{listingsutf8,newtxtt,xcolor,graphicx}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\begin{document}
\title{Übungsblatt 1}
\date{}
\maketitle
\lstset{language=Python,
  basicstyle=\ttfamily,
  keywordstyle=\bfseries\color{magenta},
  stringstyle=\color{red},
  numberstyle=\color{blue}
  breaklines=true,
  showstringspaces=false,
  backgroundcolor=\color{yellow!30},
  ndkeywordstyle=\color{yellow},
  commentstyle=\color{green},
  identifierstyle=\color{black},
  literate=
  {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
  {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
  {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
  {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
  {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
  {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
  {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
  {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
  {Ã}{{\~A}}1 {ã}{{\~a}}1 {Õ}{{\~O}}1 {õ}{{\~o}}1
  {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
  {ű}{{\H{u}}}1 {Ű}{{\H{U}}}1 {ő}{{\H{o}}}1 {Ő}{{\H{O}}}1
  {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
  {€}{{\euro}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1,
  columns=fullflexible,
  keepspaces=true
}
\makeatletter
\def\lst@outputspace{{\ifx\lst@bkgcolor\empty\color{white}\else\lst@bkgcolor\fi\lst@visiblespace}}
\makeatother
\thispagestyle{empty}
\begin{enumerate}
\item Betrachten wir folgendes Python3 Programm:
\lstinputlisting[language=Python]{../../progs/de_AT/1.py}
Tippen Sie dieses Programm unter der Website
\begin{center}\url{https://repl.it/languages/python3}\end{center}
im Fenster \textbf{main.py} ein:
\begin{center}\includegraphics[scale=0.5]{main-py-window}\end{center}
Überprüfen Sie, dass die Einzüge bei Ihnen für jede Zeile genauso aussehen, wie in der Aufgabenstellung.
Sie können den Einzug für jede Zeile am Anfang der Zeile mit der Tab-Taste und mit Shift-Tab kontrollieren.

\item Klicken Sie auf die Run-Taste. Welches Resultat ergibt sich im rechten schwarzen Fenster?

\item Erklären Sie die Bedeutung für jede Zeile des Programms in jeweils einem Satz.

\item Ändern Sie das Programm, so dass das Resultat \texttt{3 von 6} wird.
\end{enumerate}
\end{document}

### 2020-11-02 progs/de_AT/1.py

@@ -1,0 +1,7 @@
+möglich = 0
+günstig = 0
+for Augenzahl in [1,2,3,4,5,6]:
+  möglich = möglich + 1
+  if Augenzahl > 4:
+    günstig = günstig + 1
+print (f"{günstig} von {möglich}")

### 2020-11-02 progs/de_AT/2.py

@@ -1,0 +1,9 @@
+möglich = 0
+günstig = 0
+for Augenzahl_Würfel_1 in [1,2,3,4,5,6]:
+  for Augenzahl_Würfel_2 in [1,2,3,4,5,6]:
+    möglich = möglich + 1
+    if Augenzahl_Würfel_1 + Augenzahl_Würfel_2 == 10:
+      günstig = günstig + 1
+print(f"{günstig} von {möglich}")
+print(f"p = {günstig/möglich}")

### 2020-11-02 progs/de_AT/3.py

@@ -1,0 +1,18 @@
+import itertools
+# 1. Grundmenge erstellen:
+Augenzahlen = {1, 2, 3, 4, 5, 6}
+# itertools.product liefert das kartesische Produkt:
+Ω = set(itertools.product(Augenzahlen, Augenzahlen))
+print(Ω)
+# Beachte, dass Ω als Menge nicht sortiert ist.
+
+# 2. Anzahl Günstige und Mögliche bestimmen:
+günstig = 0
+möglich = len(Ω)
+for Augenzahl1, Augenzahl2 in Ω:
+  if Augenzahl1 + Augenzahl2 == 10:
+    günstig = günstig + 1
+
+# 3. Ausgabe:
+print(f"{günstig} von {möglich}")
+print(f"p = {günstig/möglich}")

### 2020-11-02 progs/de_AT/4.py

@@ -1,0 +1,14 @@
+möglich = 0
+günstig = 0
+for Münze1 in {"K","Z"}:
+  for Münze2 in {"K","Z"}:
+    for Münze3 in {"K","Z"}:
+      for Münze4 in {"K","Z"}:
+        for Münze5 in {"K","Z"}:
+          möglich = möglich + 1
+          # Eine Zeichenkette wird generiert:
+          Ausgang = Münze1 + Münze2 + Münze3 + Münze4 + Münze5 
+          if Ausgang.count("K")==2:
+            günstig = günstig + 1
+print(f"{günstig} von {möglich}")
+print(f"p = {günstig/möglich}")

### 2020-11-02 progs/de_AT/5.py

@@ -1,0 +1,22 @@
+import itertools
+# 1. Mathematische Grundmenge Ω besteht aus lauter 5-elementigen Tupeln, wobei K für Kopf und
+#    Z für Zahl steht. Jeder Buchstabe im Tupel repräsentiert den Ausgang einer geworfenen Münze.
+Ω = set(itertools.product({"K", "Z"}, repeat=5))
+print(Ω)
+
+# 2. Ω wird in eine für unser Programm brauchbarere Form transformiert. Aus den Tupeln werden
+#    Zeichenketten (strings) gemacht.
+Ω = {"".join(Ausgang) for Ausgang in Ω}
+print(Ω)
+
+# 3. Anzahl der günstigen und möglichen Ausgänge bestimmen:
+günstig = 0
+möglich = len(Ω)
+for Ausgang in Ω:
+    if Ausgang.count("K") == 2:
+        günstig = günstig + 1
+
+# 4. Ausgabe:
+print(f"{günstig} von {möglich}")
+print(f"p = {günstig/möglich}")
+

### 2020-11-02 progs/de_AT/6.py

@@ -1,0 +1,12 @@
+möglich = 0
+Erwartungswert = 0
+for Kopf1 in [0,1]:
+  for Kopf2 in [0,1]:
+    for Kopf3 in [0,1]:
+      for Kopf4 in [0,1]:
+        möglich += 1
+        x = Kopf1 + Kopf2 + Kopf3 + Kopf4
+        print(f"Ereignis={Kopf1}{Kopf2}{Kopf3}{Kopf4}, x={x}")
+        Erwartungswert = Erwartungswert + x
+Erwartungswert = Erwartungswert / möglich
+print(f"E(X)={Erwartungswert}")

### 2020-11-02 progs/de_AT/7.py

@@ -1,0 +1,14 @@
+möglich = 0
+Erwartungswert = 0
+Varianz = 0
+for Kopf1 in [0,1]:
+  for Kopf2 in [0,1]:
+    for Kopf3 in [0,1]:
+      for Kopf4 in [0,1]:
+        möglich += 1
+        x = Kopf1 + Kopf2 + Kopf3 + Kopf4
+        Erwartungswert = Erwartungswert + x
+        Varianz = Varianz + x*x
+Erwartungswert = Erwartungswert/möglich
+Varianz = Varianz/möglich - Erwartungswert*Erwartungswert
+print(f"E(X)={Erwartungswert}, V(X)={Varianz}")

### 2020-11-03 forms/de_AT/1.tex
Fix naming

@@
 \begin{center}\includegraphics[scale=0.5]{main-py-window}\end{center}
-Überprüfen Sie, dass die Einzüge bei Ihnen für jede Zeile genauso aussehen, wie in der Aufgabenstellung.
-Sie können den Einzug für jede Zeile am Anfang der Zeile mit der Tab-Taste und mit Shift-Tab kontrollieren.
+Überprüfen Sie, dass die Einrücken bei Ihnen für jede Zeile genauso aussehen, wie in der Aufgabenstellung.
+Sie können das Einrücken für jede Zeile am Anfang der Zeile mit der Tab-Taste und mit Shift-Tab kontrollieren.
 

### 2020-11-14 forms/de_AT/2.tex
Updates for the forms

--- 2020-11-02 forms/de_AT/1.tex
@@
 \begin{document}
-\title{Übungsblatt 1}
+\title{Übungsblatt 2}
 \date{}
@@
 \item Betrachten wir folgendes Python3 Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/1.py}
-Tippen Sie dieses Programm unter der Website
-\begin{center}\url{https://repl.it/languages/python3}\end{center}
-im Fenster \textbf{main.py} ein:
-\begin{center}\includegraphics[scale=0.5]{main-py-window}\end{center}
-Überprüfen Sie, dass die Einzüge bei Ihnen für jede Zeile genauso aussehen, wie in der Aufgabenstellung.
-Sie können den Einzug für jede Zeile am Anfang der Zeile mit der Tab-Taste und mit Shift-Tab kontrollieren.
-
-\item Klicken Sie auf die Run-Taste. Welches Resultat ergibt sich im rechten schwarzen Fenster?
+\lstinputlisting[language=Python]{../../progs/de_AT/2.py}
 
@@
 
-\item Ändern Sie das Programm, so dass das Resultat \texttt{3 von 6} wird.
+\item In welchem Ereignisraum werden die Ereignisse abgezählt?
+
+\item Ändern Sie das Programm, somit Sie diese Aufgabe mechanisch lösen können:
+
+\begin{quote}
+Sie würfeln dreimal hintereinander mit einem Würfel. Berechnen Sie die Wahrscheinlichkeit,
+dass die Folge der gewürfelten Augenzahlen monoton wachsend ist, unter Laplace-Annahme.
+\end{quote}
+
+Hinweis: Versuchen Sie die Aufgabe ohne Hilfsmittel zu lösen. Fall Sie kein Erfolg haben,
+können Sie die Website \url{https://www.snakify.org/de/lessons/if_then_else_conditions/}
+für einen Tipp durchlesen.
+
 \end{enumerate}

### 2020-11-17 progs/de_AT/3.py
Updates

@@ -5,6 +5,8 @@ Augenzahlen = {1, 2, 3, 4, 5, 6}
 Ω = set(itertools.product(Augenzahlen, Augenzahlen))
 print(Ω)
 # Beachte, dass Ω als Menge nicht sortiert ist.
+# Statt "set" kann auch "list" geschrieben werden:
+# dann wird die Menge schon sortiert.
 
 # 2. Anzahl Günstige und Mögliche bestimmen:
 günstig = 0

### 2020-11-17 progs/de_AT/3b.py

@@ -1,0 +1,22 @@
+import itertools
+
+def X(ω):
+    """Studenten könnten diese Funktion selbst programmieren.
+    X ist Zufallsvariable, also X: Ω → ℝ.
+    Einem Ergebnis wird eine reelle Zahl zugeordnet."""
+    return ω[0]+ω[1]
+
+Augenzahlen = {1, 2, 3, 4, 5, 6}
+# Ergebnisraum Ω ist hier eine Menge an Paaren, wobei der erste (zweite) Eintrag
+# die Augenzahl des ersten (zweiten) Würfels ist.
+Ω = list(itertools.product(Augenzahlen, Augenzahlen))
+# Ereignis E
+E = []
+for ω in Ω:
+    if X(ω) == 10:
+        E.append(ω)
+# "Anzahl der Möglichen" ist die Mächtigkeit des Ergebnisraumes
+print(f"{len(Ω)} mögliche Ergebnisse")
+# "Anzahl der Günstigen" ist die Mächtigkeit des Ereignisses
+print(f"{len(E)} günstige Ergebnisse")
+print(f"p = {len(E)/len(Ω)}")

### 2020-11-26 forms/de_AT/3.tex
Updates for 3rd form

\documentclass{article}
\usepackage{listingsutf8,newtxtt,xcolor,graphicx}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\begin{document}
\title{Übungsblatt 3}
\date{}
\maketitle
\lstset{language=Python,
  numbers=left,
  basicstyle=\ttfamily,
  keywordstyle=\bfseries\color{magenta},
  stringstyle=\color{red},
  numberstyle=\color{blue},
  breaklines=true,
  showstringspaces=false,
  backgroundcolor=\color{yellow!30},
  ndkeywordstyle=\color{magenta!70},
  commentstyle=\color{green},
  identifierstyle=\color{black},
  literate=
  {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
  {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
  {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
  {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
  {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
  {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
  {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
  {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
  {Ã}{{\~A}}1 {ã}{{\~a}}1 {Õ}{{\~O}}1 {õ}{{\~o}}1
  {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
  {ű}{{\H{u}}}1 {Ű}{{\H{U}}}1 {ő}{{\H{o}}}1 {Ő}{{\H{O}}}1
  {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
  {€}{{\euro}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1
  {Ω}{{$\Omega$}}1,
  columns=fullflexible,
  keepspaces=true
}
\makeatletter
\def\lst@outputspace{{\ifx\lst@bkgcolor\empty\color{white}\else\lst@bkgcolor\fi\lst@visiblespace}}
\makeatother
\thispagestyle{empty}
\begin{enumerate}
\item Betrachten wir folgendes Python3 Programm:
\lstinputlisting[language=Python]{../../progs/de_AT/3.py}

Erklären Sie die Bedeutung für jede Zeile des Programms in jeweils einem Satz. Hinweis:
\texttt{itertools} ist eine Sammlung von nützlichen Python-Befehlen. Dies ist notwendig
für die 5.~Zeile.

\item Nutzen Sie \texttt{list} statt \texttt{set} in der 5.~Zeile. Wie ändert sich das Resultat?

\item Zählen Sie die \texttt{Augenzahlen} in der 5.~Zeile dreimal auf und geben Sie auch die Variable
\texttt{Augenzahl3} in der 14.~Zeile nach einem weiteren Beistrich an.
Weiters, nutzen Sie Ihre Lösung für die Aufgabe 4
im 2.~Übungsblatt und erweitern Sie hier die 15.~Zeile, so dass Sie nach dieser Art eine neue Lösungsmethode
erhalten.

\item Ändern Sie Ihr Programm in der 5.~Zeile nach dieser Art:
\begin{lstlisting}[language=Python,firstnumber=5]
Ω = set(itertools.product(Augenzahlen, repeat=3))
\end{lstlisting}
Wann kann diese Schreibweise nützlich sein?

\end{enumerate}
\end{document}

### 2020-12-08 forms/de_AT/4.tex
Updates for form 4

\documentclass{article}
\usepackage{listingsutf8,newtxtt,xcolor,graphicx}
\usepackage[hyphens]{url}
\addtolength{\topmargin}{-.875in}
\addtolength{\textheight}{1.75in}
\usepackage{hyperref}
\usepackage{amsmath,amssymb}
\begin{document}
\title{Übungsblatt 4}
\date{}
\maketitle
\lstset{language=Python,
  numbers=left,
  basicstyle=\ttfamily,
  keywordstyle=\bfseries\color{magenta},
  stringstyle=\color{red},
  numberstyle=\color{blue},
  breaklines=true,
  showstringspaces=false,
  backgroundcolor=\color{yellow!30},
  ndkeywordstyle=\color{magenta!70},
  commentstyle=\color{green},
  identifierstyle=\color{black},
  literate=
  {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
  {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
  {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
  {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
  {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
  {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
  {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
  {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
  {Ã}{{\~A}}1 {ã}{{\~a}}1 {Õ}{{\~O}}1 {õ}{{\~o}}1
  {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
  {ű}{{\H{u}}}1 {Ű}{{\H{U}}}1 {ő}{{\H{o}}}1 {Ő}{{\H{O}}}1
  {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
  {€}{{\euro}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1
  {Ω}{{$\Omega$}}1
  {ω}{{$\omega$}}1
  {→}{{$\to$}}1
  {ℝ}{{$\mathbb{R}$}}1,
  columns=fullflexible,
  keepspaces=true
}
\makeatletter
\def\lst@outputspace{{\ifx\lst@bkgcolor\empty\color{white}\else\lst@bkgcolor\fi\lst@visiblespace}}
\makeatother
\thispagestyle{empty}
\begin{enumerate}
\item Betrachten wir folgendes Python3 Programm:
\lstinputlisting[language=Python]{../../progs/de_AT/4.py}

\begin{enumerate}
\item Überprüfen Sie, ob das Programm richtig funktioniert. Speichern Sie als Beweis ein Bildschirmfoto ab.
\item Erklären Sie die Unterschiede in diesem Programm und dem im Übungsblatt 3 (Aufgabe 1).
\end{enumerate}

\item Ändern Sie Ihr Programm in den 7.~und 14.~Zeilen, so dass Sie folgendes Beispiel lösen können:
\begin{quote}
Anna und Bernhard würfeln viermal hintereinander mit einem Würfel. 
Wenn die Augensumme zwischen 7 und 14 ist, gewinnt Anna, sonst Bernhard.
Welcher Spieler wird bei dem Spiel begünstigt?
\end{quote}

\item Ändern Sie die Spielregeln in der obigen Aufgabe, so dass das Spiel gerecht sei.
Überprüfen Sie Ihre Aufgabenstellung mithilfe von einem passenden Python3 Programm.

\item Wir ersetzen die Zeilen 12--15 im obigen Programm mit der Zeile
\begin{lstlisting}[language=Python,numbers=none]
E = [ω for ω in Ω if X(ω) == 10]
\end{lstlisting}
Geben Sie diese Zeile in einer mathematischen Form als eine Mengendefinition an.

\end{enumerate}
\end{document}

### 2020-12-08 progs/de_AT/3b.py

@@ -1,22 +1,0 @@
-import itertools
-
-def X(ω):
-    """Studenten könnten diese Funktion selbst programmieren.
-    X ist Zufallsvariable, also X: Ω → ℝ.
-    Einem Ergebnis wird eine reelle Zahl zugeordnet."""
-    return ω[0]+ω[1]
-
-Augenzahlen = {1, 2, 3, 4, 5, 6}
-# Ergebnisraum Ω ist hier eine Menge an Paaren, wobei der erste (zweite) Eintrag
-# die Augenzahl des ersten (zweiten) Würfels ist.
-Ω = list(itertools.product(Augenzahlen, Augenzahlen))
-# Ereignis E
-E = []
-for ω in Ω:
-    if X(ω) == 10:
-        E.append(ω)
-# "Anzahl der Möglichen" ist die Mächtigkeit des Ergebnisraumes
-print(f"{len(Ω)} mögliche Ergebnisse")
-# "Anzahl der Günstigen" ist die Mächtigkeit des Ereignisses
-print(f"{len(E)} günstige Ergebnisse")
-print(f"p = {len(E)/len(Ω)}")

### 2020-12-08 progs/de_AT/4.py

@@ -1,14 +1,18 @@
-möglich = 0
-günstig = 0
-for Münze1 in {"K","Z"}:
-  for Münze2 in {"K","Z"}:
-    for Münze3 in {"K","Z"}:
-      for Münze4 in {"K","Z"}:
-        for Münze5 in {"K","Z"}:
-          möglich = möglich + 1
-          # Eine Zeichenkette wird generiert:
-          Ausgang = Münze1 + Münze2 + Münze3 + Münze4 + Münze5 
-          if Ausgang.count("K")==2:
-            günstig = günstig + 1
-print(f"{günstig} von {möglich}")
-print(f"p = {günstig/möglich}")
+import itertools
+
+def X(ω):
+    """Studenten können diese Funktion selbst programmieren.
+    X ist Zufallsvariable, also X: Ω → ℝ.
+    Einem Ergebnis wird eine reelle Zahl zugeordnet."""
+    return ω[0]+ω[1]
+
+Augenzahlen = {1, 2, 3, 4, 5, 6}
+Ω = list(itertools.product(Augenzahlen, Augenzahlen))
+# Ereignis E
+E = []
+for ω in Ω:
+    if X(ω) == 10:
+        E.append(ω)
+print(f"{len(Ω)} mögliche Ergebnisse")
+print(f"{len(E)} günstige Ergebnisse")
+print(f"p = {len(E)/len(Ω)}")

### 2020-12-08 progs/de_AT/5.py

@@ -1,22 +1,14 @@
-import itertools
-# 1. Mathematische Grundmenge Ω besteht aus lauter 5-elementigen Tupeln, wobei K für Kopf und
-#    Z für Zahl steht. Jeder Buchstabe im Tupel repräsentiert den Ausgang einer geworfenen Münze.
-Ω = set(itertools.product({"K", "Z"}, repeat=5))
-print(Ω)
-
-# 2. Ω wird in eine für unser Programm brauchbarere Form transformiert. Aus den Tupeln werden
-#    Zeichenketten (strings) gemacht.
-Ω = {"".join(Ausgang) for Ausgang in Ω}
-print(Ω)
-
-# 3. Anzahl der günstigen und möglichen Ausgänge bestimmen:
+möglich = 0
 günstig = 0
-möglich = len(Ω)
-for Ausgang in Ω:
-    if Ausgang.count("K") == 2:
-        günstig = günstig + 1
-
-# 4. Ausgabe:
+for Münze1 in {"K","Z"}:
+  for Münze2 in {"K","Z"}:
+    for Münze3 in {"K","Z"}:
+      for Münze4 in {"K","Z"}:
+        for Münze5 in {"K","Z"}:
+          möglich = möglich + 1
+          # Eine Zeichenkette wird generiert:
+          Ausgang = Münze1 + Münze2 + Münze3 + Münze4 + Münze5 
+          if Ausgang.count("K")==2:
+            günstig = günstig + 1
 print(f"{günstig} von {möglich}")
 print(f"p = {günstig/möglich}")
-

### 2020-12-08 progs/de_AT/6.py

@@ -1,12 +1,22 @@
-möglich = 0
-Erwartungswert = 0
-for Kopf1 in [0,1]:
-  for Kopf2 in [0,1]:
-    for Kopf3 in [0,1]:
-      for Kopf4 in [0,1]:
-        möglich += 1
-        x = Kopf1 + Kopf2 + Kopf3 + Kopf4
-        print(f"Ereignis={Kopf1}{Kopf2}{Kopf3}{Kopf4}, x={x}")
-        Erwartungswert = Erwartungswert + x
-Erwartungswert = Erwartungswert / möglich
-print(f"E(X)={Erwartungswert}")
+import itertools
+# 1. Mathematische Grundmenge Ω besteht aus lauter 5-elementigen Tupeln, wobei K für Kopf und
+#    Z für Zahl steht. Jeder Buchstabe im Tupel repräsentiert den Ausgang einer geworfenen Münze.
+Ω = set(itertools.product({"K", "Z"}, repeat=5))
+print(Ω)
+
+# 2. Ω wird in eine für unser Programm brauchbarere Form transformiert. Aus den Tupeln werden
+#    Zeichenketten (strings) gemacht.
+Ω = {"".join(Ausgang) for Ausgang in Ω}
+print(Ω)
+
+# 3. Anzahl der günstigen und möglichen Ausgänge bestimmen:
+günstig = 0
+möglich = len(Ω)
+for Ausgang in Ω:
+    if Ausgang.count("K") == 2:
+        günstig = günstig + 1
+
+# 4. Ausgabe:
+print(f"{günstig} von {möglich}")
+print(f"p = {günstig/möglich}")
+

### 2020-12-08 progs/de_AT/8.py

@@ -1,12 +1,14 @@
 möglich = 0
 Erwartungswert = 0
+Varianz = 0
 for Kopf1 in [0,1]:
   for Kopf2 in [0,1]:
     for Kopf3 in [0,1]:
       for Kopf4 in [0,1]:
         möglich += 1
         x = Kopf1 + Kopf2 + Kopf3 + Kopf4
-        print(f"Ereignis={Kopf1}{Kopf2}{Kopf3}{Kopf4}, x={x}")
         Erwartungswert = Erwartungswert + x
-Erwartungswert = Erwartungswert / möglich
-print(f"E(X)={Erwartungswert}")
+        Varianz = Varianz + x*x
+Erwartungswert = Erwartungswert/möglich
+Varianz = Varianz/möglich - Erwartungswert*Erwartungswert
+print(f"E(X)={Erwartungswert}, V(X)={Varianz}")

### 2020-12-17 forms/de_AT/5.tex
Update

--- 2020-12-08 forms/de_AT/4.tex
@@
 \begin{document}
-\title{Übungsblatt 4}
+\title{Übungsblatt 5}
 \date{}
@@
 \item Betrachten wir folgendes Python3 Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/4.py}
+\lstinputlisting[language=Python]{../../progs/de_AT/5a.py}
 
 \begin{enumerate}
-\item Überprüfen Sie, ob das Programm richtig funktioniert. Speichern Sie als Beweis ein Bildschirmfoto ab.
-\item Erklären Sie die Unterschiede in diesem Programm und dem im Übungsblatt 3 (Aufgabe 1).
+\item Finden Sie eine passende Textaufgabe aus der Wahrscheinlichkeits\-rechnung, welche
+von diesem Programm gelöst wird.
+\item Geben Sie eine kurze mathematische Begründung an, warum die Antwort des Programms $\binom{5}{2}/2^5$ ist.
 \end{enumerate}
 
-\item Ändern Sie Ihr Programm in den 7.~und 14.~Zeilen, so dass Sie folgendes Beispiel lösen können:
-\begin{quote}
-Anna und Bernhard würfeln viermal hintereinander mit einem Würfel. 
-Wenn die Augensumme zwischen 7 und 14 ist, gewinnt Anna, sonst Bernhard.
-Welcher Spieler wird bei dem Spiel begünstigt?
-\end{quote}
+\item Betrachten wir folgendes Python3 Programm:
+\lstinputlisting[language=Python]{../../progs/de_AT/5b.py}
 
-\item Ändern Sie die Spielregeln in der obigen Aufgabe, so dass das Spiel gerecht sei.
-Überprüfen Sie Ihre Aufgabenstellung mithilfe von einem passenden Python3 Programm.
-
-\item Wir ersetzen die Zeilen 12--15 im obigen Programm mit der Zeile
-\begin{lstlisting}[language=Python,numbers=none]
-E = [ω for ω in Ω if X(ω) == 10]
-\end{lstlisting}
-Geben Sie diese Zeile in einer mathematischen Form als eine Mengendefinition an.
+\begin{enumerate}
+\item Geben Sie die Zufallsvariable $X$ an, wessen Erwartungswert von diesem zweiten Programm ausgerechnet wird.
+\item Geben Sie eine kurze mathematische Begründung an, warum der Erwartungswert $2.5$ ist.
+Verallgemeinern Sie dieses Resultat auch für unterschiedliche Werte bei der Einstellung
+\texttt{repeat=5} in der 2.~Zeile des Programms.
+\item Erweitern Sie dieses zweite Programm, so dass auch die Varianz von $X$ berechnet wird.
+\end{enumerate}
 

### 2020-12-17 progs/de_AT/5.py

@@ -7,7 +7,7 @@ for Münze1 in {"K","Z"}:
         for Münze5 in {"K","Z"}:
           möglich = möglich + 1
           # Eine Zeichenkette wird generiert:
-          Ausgang = Münze1 + Münze2 + Münze3 + Münze4 + Münze5 
+          Ausgang = Münze1 + Münze2 + Münze3 + Münze4 + Münze5
           if Ausgang.count("K")==2:
             günstig = günstig + 1
 print(f"{günstig} von {möglich}")

### 2020-12-17 progs/de_AT/5a.py

@@ -1,0 +1,14 @@
+import itertools
+Ω = list(itertools.product({"K", "Z"}, repeat=5))
+
+# Ω wird in eine für unser Programm
+# brauchbarere Form transformiert.
+# Aus den Tupeln werden Zeichenketten (strings) gemacht.
+Ω = {"".join(Ausgang) for Ausgang in Ω}
+print(f"Ω={Ω}")
+
+E = [ω for ω in Ω if ω.count("K") == 2]
+print(f"E={E}")
+
+print(f"{len(E)} von {len(Ω)}")
+print(f"p = {len(E)/len(Ω)}")

### 2020-12-17 progs/de_AT/5b.py

@@ -1,0 +1,10 @@
+import itertools
+Ω = set(itertools.product({"K", "Z"}, repeat=5))
+Ω = {"".join(Ausgang) for Ausgang in Ω}
+
+Erwartungswert = 0
+for ω in Ω:
+  Erwartungswert = Erwartungswert + ω.count("K")
+Erwartungswert = Erwartungswert / len(Ω)
+
+print(f"Erwartungswert = {Erwartungswert}")

### 2020-12-17 progs/de_AT/6.py

@@ -1,22 +1,18 @@
 import itertools
-# 1. Mathematische Grundmenge Ω besteht aus lauter 5-elementigen Tupeln, wobei K für Kopf und
-#    Z für Zahl steht. Jeder Buchstabe im Tupel repräsentiert den Ausgang einer geworfenen Münze.
 Ω = set(itertools.product({"K", "Z"}, repeat=5))
 print(Ω)
 
-# 2. Ω wird in eine für unser Programm brauchbarere Form transformiert. Aus den Tupeln werden
-#    Zeichenketten (strings) gemacht.
+# Ω wird in eine für unser Programm brauchbarere Form transformiert.
+# Aus den Tupeln werden Zeichenketten (strings) gemacht.
 Ω = {"".join(Ausgang) for Ausgang in Ω}
 print(Ω)
 
-# 3. Anzahl der günstigen und möglichen Ausgänge bestimmen:
 günstig = 0
 möglich = len(Ω)
 for Ausgang in Ω:
     if Ausgang.count("K") == 2:
         günstig = günstig + 1
 
-# 4. Ausgabe:
 print(f"{günstig} von {möglich}")
 print(f"p = {günstig/möglich}")
 

### 2020-12-17 progs/de_AT/5a.py
Update (applying Alex's suggestions)

@@ -1,14 +1,7 @@
 import itertools
 Ω = list(itertools.product({"K", "Z"}, repeat=5))
-
-# Ω wird in eine für unser Programm
-# brauchbarere Form transformiert.
-# Aus den Tupeln werden Zeichenketten (strings) gemacht.
-Ω = {"".join(Ausgang) for Ausgang in Ω}
 print(f"Ω={Ω}")
-
 E = [ω for ω in Ω if ω.count("K") == 2]
 print(f"E={E}")
-
 print(f"{len(E)} von {len(Ω)}")
 print(f"p = {len(E)/len(Ω)}")

### 2020-12-17 progs/de_AT/5b.py

@@ -1,10 +1,7 @@
 import itertools
 Ω = set(itertools.product({"K", "Z"}, repeat=5))
-Ω = {"".join(Ausgang) for Ausgang in Ω}
-
 Erwartungswert = 0
 for ω in Ω:
   Erwartungswert = Erwartungswert + ω.count("K")
 Erwartungswert = Erwartungswert / len(Ω)
-
 print(f"Erwartungswert = {Erwartungswert}")

### 2021-01-22 forms/de_AT/6.tex
Add new forms

--- 2020-12-17 forms/de_AT/5.tex
@@
 \usepackage{listingsutf8,newtxtt,xcolor,graphicx}
-\usepackage[hyphens]{url}
+\usepackage[hyphens]{xurl}
 \addtolength{\topmargin}{-.875in}
@@
 \usepackage{amsmath,amssymb}
+\usepackage{geometry}
 \begin{document}
-\title{Übungsblatt 5}
+\title{Übungsblatt 6}
 \date{}
@@
 \begin{enumerate}
-\item Betrachten wir folgendes Python3 Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/5a.py}
+\item Die folgenden Python3 Programme berechnen dieselben Resultate:
+\lstinputlisting[language=Python]{../../progs/de_AT/6a.py}
+\lstinputlisting[language=Python]{../../progs/de_AT/6b.py}
 
 \begin{enumerate}
-\item Finden Sie eine passende Textaufgabe aus der Wahrscheinlichkeits\-rechnung, welche
-von diesem Programm gelöst wird.
-\item Geben Sie eine kurze mathematische Begründung an, warum die Antwort des Programms $\binom{5}{2}/2^5$ ist.
+\item Überprüfen Sie, dass die Resultate übereinstimmen, auch für 6 oder 7 Würfe.
+\item Erklären Sie, wie die Programme funktionieren. Welche Unterschiede finden Sie zwischen denen?
 \end{enumerate}
 
-\item Betrachten wir folgendes Python3 Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/5b.py}
+\eject
+
+\item Betrachten wir folgendes Programm:
+\lstinputlisting[language=Python]{../../progs/de_AT/6c.py}
 
 \begin{enumerate}
-\item Geben Sie die Zufallsvariable $X$ an, wessen Erwartungswert von diesem zweiten Programm ausgerechnet wird.
-\item Geben Sie eine kurze mathematische Begründung an, warum der Erwartungswert $2.5$ ist.
-Verallgemeinern Sie dieses Resultat auch für unterschiedliche Werte bei der Einstellung
-\texttt{repeat=5} in der 2.~Zeile des Programms.
-\item Erweitern Sie dieses zweite Programm, so dass auch die Varianz von $X$ berechnet wird.
+\item Welche Berechnung wird vom Programm durchgeführt?
+\item Betrachten wir folgende Änderungen:
+\begin{lstlisting}[language=Python,firstnumber=2]
+Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=4))
+mögliche_Werte = range(30)
+\end{lstlisting}
+\begin{lstlisting}[language=Python,firstnumber=8]
+    Anzahl_Köpfe[sum(ω)] += 1
+\end{lstlisting}
+Erklären Sie das Resultat von diesem geänderten Programm!
+\item Wir versuchen nun ein graphisches Diagramm zu erstellen. Deswegen
+betrachten wir folgende Änderungen (statt der 10.~Zeile):
+\begin{lstlisting}[language=Python,firstnumber=10]
+for w in zip(*(mögliche_Werte, Wahrscheinlichkeiten)):
+    print(f"{w[0]},{w[1]}")
+\end{lstlisting}
+Kopieren Sie den Output dieses Programms in ein Tabellenkalkulationsprogramm ein
+und erstellen Sie ein XY-Diagramm, sodass die Wahrscheinlichkeitsverteilung
+graphisch dargestellt wird! Tipp: Die Sprache der Tabelle muss auf Englisch
+geändert werden, sodass der Dezimalpunkt von Python richtig interpretiert wird.
+Hinweis: Google Tabellen kann die Beistriche auch als Trennungszeichen interpretieren,
+siehe \url{http://apps-experts.de/2016/03/neue-funktion-text-in-spalten-fuer-google-tabellen/}.
+\item Ergänzen Sie das Programm mit den nächsten zwei Zeilen:
+\begin{lstlisting}[language=Python,firstnumber=12]
+Verteilung = [sum(Wahrscheinlichkeiten[:h]) for h in mögliche_Werte]
+print(list(zip(*(mögliche_Werte, Verteilung))))
+\end{lstlisting}
+Erklären Sie, welche Berechnung hier durchgeführt wird.
 \end{enumerate}
-
 \end{enumerate}

### 2021-01-22 forms/de_AT/7.tex
Add new forms

--- 2020-12-17 forms/de_AT/5.tex
@@
 \usepackage{amsmath,amssymb}
+\usepackage{geometry}
 \begin{document}
-\title{Übungsblatt 5}
+\title{Übungsblatt 7}
 \date{}
@@
 \thispagestyle{empty}
-\begin{enumerate}
-\item Betrachten wir folgendes Python3 Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/5a.py}
+
+\noindent Betrachten wir folgendes Programm:
+\lstinputlisting[language=Python]{../../progs/de_AT/7.py}
 
 \begin{enumerate}
-\item Finden Sie eine passende Textaufgabe aus der Wahrscheinlichkeits\-rechnung, welche
-von diesem Programm gelöst wird.
-\item Geben Sie eine kurze mathematische Begründung an, warum die Antwort des Programms $\binom{5}{2}/2^5$ ist.
-\end{enumerate}
-
-\item Betrachten wir folgendes Python3 Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/5b.py}
-
+\item Welche Resultate werden vom Programm berechnet?
+\item Erstellen Sie ein Liniendiagramm, welches die Resultate grafisch darstellt.
+\item Erklären Sie mit möglichst präzisen mathematischen Begriffen, wie das Liniendiagramm interpretiert werden kann.
+Wichtige Hinweise:
 \begin{enumerate}
-\item Geben Sie die Zufallsvariable $X$ an, wessen Erwartungswert von diesem zweiten Programm ausgerechnet wird.
-\item Geben Sie eine kurze mathematische Begründung an, warum der Erwartungswert $2.5$ ist.
-Verallgemeinern Sie dieses Resultat auch für unterschiedliche Werte bei der Einstellung
-\texttt{repeat=5} in der 2.~Zeile des Programms.
-\item Erweitern Sie dieses zweite Programm, so dass auch die Varianz von $X$ berechnet wird.
+\item Lesen Sie so viele Eigenschaften ab, wie es möglich ist.
+\item Verallgemeinern Sie die Resultate.
+\item Nutzen Sie nur solche Begriffe, die in der Schule (Unterstufe oder Oberstufe) erklärt werden können.
+(Wir nehmen an, dass folgende Begriffe schon bekannt sind: Folge, Grenz\-wert, Funktion,
+Dichtefunktion, Erwartungswert, Varianz, Streuung, Normalverteilung.)
 \end{enumerate}

### 2021-01-22 progs/de_AT/6a.py

@@ -1,0 +1,12 @@
+import itertools
+Ω = set(itertools.product({"K", "Z"}, repeat=5))
+E = 0
+for ω in Ω:
+  E = E + ω.count("K")
+E = E / len(Ω)
+V = 0
+for ω in Ω:
+  V = V + (ω.count("K")-E)**2
+V = V / len(Ω)
+print(f"Erwartungswert = {E}")
+print(f"Varianz = {V}")

### 2021-01-22 progs/de_AT/6b.py

@@ -1,0 +1,12 @@
+import itertools
+Ω = set(itertools.product({"K", "Z"}, repeat=5))
+E = 0
+V = 0
+for ω in Ω:
+  x = ω.count("K")
+  E = E + x
+  V = V + x*x
+E = E / len(Ω)
+V = V / len(Ω) - E*E
+print(f"Erwartungswert = {E}")
+print(f"Varianz = {V}")

### 2021-01-22 progs/de_AT/6c.py

@@ -1,0 +1,10 @@
+import itertools
+Ω = set(itertools.product({"K", "Z"}, repeat=5))
+mögliche_Werte = [0, 1, 2, 3, 4, 5]
+# Zu Beginn hat jeder mögliche Wert der Zufallsvariablen
+# eine Häufigkeit von 0.
+Anzahl_Köpfe = [0]*len(mögliche_Werte)
+for ω in Ω:
+    Anzahl_Köpfe[ω.count("K")] += 1
+Wahrscheinlichkeiten = [h/len(Ω) for h in Anzahl_Köpfe]
+print(list(zip(*(mögliche_Werte, Wahrscheinlichkeiten))))

### 2021-01-22 progs/de_AT/7.py

@@ -1,12 +1,9 @@
-möglich = 0
-Erwartungswert = 0
-for Kopf1 in [0,1]:
-  for Kopf2 in [0,1]:
-    for Kopf3 in [0,1]:
-      for Kopf4 in [0,1]:
-        möglich += 1
-        x = Kopf1 + Kopf2 + Kopf3 + Kopf4
-        print(f"Ereignis={Kopf1}{Kopf2}{Kopf3}{Kopf4}, x={x}")
-        Erwartungswert = Erwartungswert + x
-Erwartungswert = Erwartungswert / möglich
-print(f"E(X)={Erwartungswert}")
+import itertools
+max_Würfe = 4
+Anzahl_Köpfe = [[0] * (max_Würfe * 7)] * (max_Würfe + 1)
+for W in range(1, max_Würfe + 1):
+    Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
+    for ω in Ω:
+        Anzahl_Köpfe[W][sum(ω)] += 1
+    Wahrscheinlichkeiten = [h/len(Ω) for h in Anzahl_Köpfe[W]]
+    print(Wahrscheinlichkeiten)

### 2021-01-23 forms/de_AT/7.tex
Update form 7

@@
 \usepackage[hyphens]{xurl}
-\addtolength{\topmargin}{-.875in}
-\addtolength{\textheight}{1.75in}
 \usepackage{hyperref}
@@
 \usepackage{geometry}
+\addtolength{\topmargin}{-.875in}
+\addtolength{\textheight}{1.75in}
 \begin{document}
@@
 \noindent Betrachten wir folgendes Programm:
-\lstinputlisting[language=Python]{../../progs/de_AT/7.py}
+\lstinputlisting[language=Python]{../../progs/de_AT/7a.py}
 
@@
 \item Welche Resultate werden vom Programm berechnet?
-\item Erstellen Sie ein Liniendiagramm, welches die Resultate grafisch darstellt.
-\item Erklären Sie mit möglichst präzisen mathematischen Begriffen, wie das Liniendiagramm interpretiert werden kann.
+\item Erstellen Sie Liniendiagramme, welche die Resultate grafisch darstellen. (Für jede
+Dichtefunktion sollten Sie eine einzige Kurve erstellen und alle Kurven indemselben Graphen
+dar\-stellen.) Tipp: Falls Sie die Website \url{cocalc.com} nutzen und ein Sage Worksheet
+erstellen, können Sie alle Liniendiagramme automatisch darstellen:
+\lstinputlisting[language=Python]{../../progs/de_AT/7b.py}
+Eventuell können Sie statt der 10.~Zeile folgenden Befehl nutzen:
+\begin{lstlisting}[language=Python,firstnumber=10]
+    p.bar(range(1, max_Würfe * 6 + 1), Wahrscheinlichkeiten[1:], alpha=0.5)
+\end{lstlisting}
+Welchen grafischen Unterschied gibt es zwischen den zwei Darstellungen?
+\item Erklären Sie mit möglichst präzisen mathematischen Begriffen, wie die Liniendiagramme interpretiert werden können.
 Wichtige Hinweise:

### 2021-01-23 progs/de_AT/7a.py

@@ -1,9 +1,9 @@
 import itertools
 max_Würfe = 4
-Anzahl_Köpfe = [[0] * (max_Würfe * 7)] * (max_Würfe + 1)
 for W in range(1, max_Würfe + 1):
     Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
+    Augensummen = [0] * (max_Würfe * 6 + 1)
     for ω in Ω:
-        Anzahl_Köpfe[W][sum(ω)] += 1
-    Wahrscheinlichkeiten = [h/len(Ω) for h in Anzahl_Köpfe[W]]
+        Augensummen[sum(ω)] += 1
+    Wahrscheinlichkeiten = [h/len(Ω) for h in Augensummen]
     print(Wahrscheinlichkeiten)

### 2021-01-23 progs/de_AT/7b.py

@@ -1,0 +1,11 @@
+import itertools
+import matplotlib.pyplot as p
+max_Würfe = 4
+for W in range(1, max_Würfe + 1):
+    Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
+    Augensummen = [0] * (max_Würfe * 6 + 1)
+    for ω in Ω:
+        Augensummen[sum(ω)] += 1
+    Wahrscheinlichkeiten = [h / len(Ω) for h in Augensummen]
+    p.plot(range(1, max_Würfe * 6 + 1), Wahrscheinlichkeiten[1:])
+p.show()

### 2021-01-23 forms/de_AT/7.tex
Fix typo

@@
 \item Erstellen Sie Liniendiagramme, welche die Resultate grafisch darstellen. (Für jede
-Dichtefunktion sollten Sie eine einzige Kurve erstellen und alle Kurven indemselben Graphen
+Dichtefunktion sollten Sie eine einzige Kurve erstellen und alle Kurven in demselben Graphen
 dar\-stellen.) Tipp: Falls Sie die Website \url{cocalc.com} nutzen und ein Sage Worksheet

### 2021-02-16 paper/article.tex
Add first version of article

% interacttfssample.tex
% v1.05 - August 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
%\setlength\bibindent{2em}% To increase hanging indent in bibliography when line spacing is doubled

\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

\begin{document}

\articletype{ARTICLE TEMPLATE}% Specify the article type or omit as appropriate

\title{Towards understanding the central limit theorem by learning Python basics}

\author{
\name{Zolt\'an Kov\'acs\textsuperscript{a}\thanks{CONTACT Z. Kov\'acs. Email: zoltan@geogebra.org}
and Alexander Thaller\textsuperscript{b}}
\affil{\textsuperscript{a}
The Private University College of Education of the Diocese of Linz,
Salesianumweg 3, Linz, Austria;
\textsuperscript{b}Linz School of Education,
Altenberger Stra\ss e 54, Linz, Austria}
}

\maketitle

\begin{abstract}

We report on a first experiment about an email based course that connects learning
Python basics and introductory probability theory. In the experiment 7 short sequences
of homework were sent out to prospective mathematics teachers who did not have
any programming background formerly, but already had some minor knowledge on probability theory.
The experiment was about to decide if learning basics of programming can promote
understanding main concepts of probability theory.
\end{abstract}

\begin{keywords}
Python; programming; probability theory; central limit theorem
\end{keywords}

\section{Introduction}

We, the authors, have been spending a reasonable time with private tutoring in many fields
of mathematics---we are teachers. And, we agree on that probability theory is maybe
the most difficult field of mathematics---if it is about \textit{understanding}. For us, rough explanation
of statistical relationships seems to be easier than explaining the delicate issues
of properties of binomial coefficients, infinite sums and convergence. Indeed, in some
sense, statistics is much easier to explain---one needs to run virtual experiments
in a web browser % add links to Steve Phelps' https://www.geogebra.org/m/bxytp4hq, Ice Cream
% https://www.geogebra.org/m/qsjbqjfe, Broken Stick
% https://www.geogebra.org/m/bu5sxbn2, Fair Coin
% + Andreas Lindner's collections: https://www.geogebra.org/m/AytaSakt, https://www.geogebra.org/m/qXPnyCTZ
and this is well supported for several years, among others, in the online version of GeoGebra.

Connecting pure mathematics and real life or virtual experiments, can be, to our experience,
extremely hard. Probability is an important item of the curriculum at secondary level,
thus it is also an important item in the teacher training. But, to our experience,
most prospective teachers \textit{never} understand the basics beyond introductory level.
Our conjecture is that the main problem lies in students' verification if theory indeed meets practice,
so, at the end of the day, a very uncertain knowledge can be observed in the students' perception. As a consequence,
in most cases, not just yesterday's and today's mathematics teachers cannot connect
theory and practice, but their students either. Roughly speaking: everybody talks about probability and
statistics, but nobody understands anything behind the scenes.

In our paper we propose a radical way, or at least, to extend the classic way how
basics of elementary probability theory should be taught. It is \textit{programming}.
Our conjecture is that understanding will be significantly improved if the main concepts
of probability theory are supported with simple computer programs. The ouput for their flexible input
should be immediately checked, eventually in a web browser, and by changing the input parameters
to a higher number, the results can be quickly generalized. We emphasize that computers
can take on the high number of computatations from human---and this is what we exactly want.
A quite simple \textit{sample space} can contain \textit{a lot} of elements, even millions or more, and it cannot
be expected that non-experts can count their elements without deeper knowledge. In fact,
experts in probability are usually experts in combinatorics as well. To reach a reasonable
level of understanding of the main concepts in probability, by using just paper and pencil,
one needs to have a very strong and \textit{safe} background in combinatorics.

In this paper we demonstrate our proposal by explaining an email based course, sent out
to 4 prospective mathematics teachers, allowing them an arbitrary time to solve the problems being set.
The students claimed that they did not have any deeper knowledge in programming formerly, but all of
them already studied probability theory at university level. Their knowledge, however, was
not yet checked via examination. The course consisted of 7 sets of homework assignments
and took place between November 2020 and January 2021, at the Johannes Kepler University of Linz, Austria.

\section{Preparations}

\section*{Acknowledgement(s)}

\begin{thebibliography}{99}

\end{thebibliography}

\end{document}

### 2021-02-16 progs/de_AT/10.py
Cleanup: remove unnecessary Python files

@@ -1,12 +1,0 @@
-möglich = 0
-Erwartungswert = 0
-for Kopf1 in [0,1]:
-  for Kopf2 in [0,1]:
-    for Kopf3 in [0,1]:
-      for Kopf4 in [0,1]:
-        möglich += 1
-        x = Kopf1 + Kopf2 + Kopf3 + Kopf4
-        print(f"Ereignis={Kopf1}{Kopf2}{Kopf3}{Kopf4}, x={x}")
-        Erwartungswert = Erwartungswert + x
-Erwartungswert = Erwartungswert / möglich
-print(f"E(X)={Erwartungswert}")

### 2021-02-16 progs/de_AT/8.py

@@ -1,14 +1,0 @@
-möglich = 0
-Erwartungswert = 0
-Varianz = 0
-for Kopf1 in [0,1]:
-  for Kopf2 in [0,1]:
-    for Kopf3 in [0,1]:
-      for Kopf4 in [0,1]:
-        möglich += 1
-        x = Kopf1 + Kopf2 + Kopf3 + Kopf4
-        Erwartungswert = Erwartungswert + x
-        Varianz = Varianz + x*x
-Erwartungswert = Erwartungswert/möglich
-Varianz = Varianz/möglich - Erwartungswert*Erwartungswert
-print(f"E(X)={Erwartungswert}, V(X)={Varianz}")

### 2021-02-16 progs/de_AT/9.py

@@ -1,18 +1,0 @@
-import itertools
-Ω = set(itertools.product({"K", "Z"}, repeat=5))
-print(Ω)
-
-# Ω wird in eine für unser Programm brauchbarere Form transformiert.
-# Aus den Tupeln werden Zeichenketten (strings) gemacht.
-Ω = {"".join(Ausgang) for Ausgang in Ω}
-print(Ω)
-
-günstig = 0
-möglich = len(Ω)
-for Ausgang in Ω:
-    if Ausgang.count("K") == 2:
-        günstig = günstig + 1
-
-print(f"{günstig} von {möglich}")
-print(f"p = {günstig/möglich}")
-

### 2021-02-17 paper/article.tex
Add new section

@@
 \newtheorem{notation}{Notation}
+
+% Own definitions (Zoltan)
+\usepackage{listingsutf8,newtxtt,xcolor,graphicx,xurl}
+\lstset{language=Python,
+  basicstyle=\ttfamily,
+  keywordstyle=\bfseries\color{magenta},
+  stringstyle=\color{red},
+  numberstyle=\color{blue}
+  breaklines=true,
+  showstringspaces=false,
+  backgroundcolor=\color{yellow!30},
+  ndkeywordstyle=\color{yellow},
+  commentstyle=\color{green},
+  identifierstyle=\color{black},
+  literate=
+  {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
+  {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
+  {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
+  {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
+  {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
+  {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
+  {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
+  {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
+  {Ã}{{\~A}}1 {ã}{{\~a}}1 {Õ}{{\~O}}1 {õ}{{\~o}}1
+  {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
+  {ű}{{\H{u}}}1 {Ű}{{\H{U}}}1 {ő}{{\H{o}}}1 {Ő}{{\H{O}}}1
+  {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
+  {€}{{\euro}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
+  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1,
+  columns=fullflexible,
+  keepspaces=true
+}
+\makeatletter
+\def\lst@outputspace{{\ifx\lst@bkgcolor\empty\color{white}\else\lst@bkgcolor\fi\lst@visiblespace}}
+\makeatother
 
@@
 
+First author hold a series of supplementary lectures on probability theory two years ago, where second author
+took part a student and illustrated several homework assignments by using Python programs.
+Some of his illustrations were a bit complicated for beginners, so the idea came that
+simplified versions of such programs could be helpful for many students.
+But, instead of giving a problem in probability theory and asking for a solution, in this
+project we reversed the logic: We revealed the Python code in advance, and wanted to know
+what problem will be solved with its help.
+
+%\begin{figure}
+%\lstinputlisting[language=Python]{../progs/de_AT/1.py}
+%\caption{The first program in its original form (in German)}
+%\label{1.py}
+%\end{figure}
+
+The first example is shown in Fig.~\ref{main-py-window}. We kept the original version to highlight
+the possibility to write special characters in the Python code (we used Python version 3),
+including accented German letters. In the first set of problems each student had to try
+the code on his or her own on the website \url{https://repl.it/languages/python3},
+by paying attention on the correct indentation as well. We instructed all students with an
+explanatory screenshot how the program should be started and the output obtained.
+
+\begin{figure}
+\begin{center}\includegraphics[scale=0.5]{../forms/de_AT/main-py-window}
+\caption{An explanatory screenshot to help the students' first steps}
+\label{main-py-window}
+\end{center}
+\end{figure}
+
+In this introductory set the first question was to make the program work by running it.
+To give a somewhat creative assignment we also asked how the output 3 can be changed to
+the number 6. The outcome of this puzzle will be analyzed later in Section \ref{problem-sets}.
+
+In fact, we created a couple of other programs in advance, but always waited for
+all solutions turned in by the students before finalizing the next set of problems.
+We \textit{did not} give any feedback on their solutions. Instead, if we felt that
+some difficulties arose, then we tried to cover them in the next set of problems.
+The reason behind this was to test if it is possible to create a set of problems
+\textit{in advance} for a possible \textit{static} learning path. Even if our experiment
+was not fully static, for a future project now we already have a static set of materials.
+
+Here we emphasize that easy access to an online version of a Python interpreter
+played a significant role in our project. We did not want to ask the students
+to install anything on their computer. Instead, we tried to minimize the gap
+between theory and practice. In fact, \url{repl.it} is not the only website
+allowing the user to try his or her code freely---we were just
+satisfied with its simple user interface. Later, to obtain graphical output in an easy way, we also
+introduced how Sage worksheets at \url{cocalc.com} can be accessed---in this latter case,
+however, accented letters did not look elegant.
+
+Importantly, we simplified and minimized the code we used during the email course.
+This pre-process required several discussions among the authors to avoid
+the technical difficulties of the programming language, but still allow
+obtaining simple outputs. We had to balance between readability, simplicity,
+elegance and speed. We learned that Python itself could be improved to
+have simpler constructions than the \texttt{for i in range(...)} loop for iteration.
+At the end of day we were still satisfied with all of the code and the students'
+solutions confirm that our choices were acceptable.
+
+As an example of a successful balance, we mention that we heavily used Python's support
+for Unicode variable names.
+We denoted the sample space by $\Omega$ and a single outcome by $\omega$ since
+these notations were already well-known for the students. On the other hand,
+typing these Greek characters is not straightforward: one may need to copy-paste
+them since they are not available on a German keyboard.
+
+\section{The problem sets}
+\label{problem-sets}
+
 \section*{Acknowledgement(s)}

### 2021-02-17 paper/article.tex
Minor update on the text

@@
 the code on his or her own on the website \url{https://repl.it/languages/python3},
-by paying attention on the correct indentation as well. We instructed all students with an
+by paying attention on the correct indentation as well---the students, however, did not have
+to type the code but just copy it from the email into the programming window.
+We also instructed all students with an
 explanatory screenshot how the program should be started and the output obtained.

### 2021-02-18 paper/article.tex
Add information on the first two sheets

@@
 \lstset{language=Python,
+  numbers=left,
   basicstyle=\ttfamily,
@@
   stringstyle=\color{red},
-  numberstyle=\color{blue}
+  numberstyle=\color{blue},
   breaklines=true,
@@
   backgroundcolor=\color{yellow!30},
-  ndkeywordstyle=\color{yellow},
+  ndkeywordstyle=\color{magenta!70},
   commentstyle=\color{green},
@@
 them already studied probability theory at university level. Their knowledge, however, was
-not yet checked via examination. The course consisted of 7 sets of homework assignments
+not yet checked via examination. The course consisted of 7 sheets of homework assignments
 and took place between November 2020 and January 2021, at the Johannes Kepler University of Linz, Austria.
@@
 \section{Preparations}
+\label{preps}
 
@@
 the possibility to write special characters in the Python code (we used Python version 3),
-including accented German letters. In the first set of problems each student had to try
+including accented German letters. In the first sheet of problems each student had to try
 the code on his or her own on the website \url{https://repl.it/languages/python3},
@@
 
-In this introductory set the first question was to make the program work by running it.
-To give a somewhat creative assignment we also asked how the output 3 can be changed to
-the number 6. The outcome of this puzzle will be analyzed later in Section \ref{problem-sets}.
+In this introductory sheet the first question was to make the program work by running it,
+and then explain its meaning line by line.
+To give a somewhat creative assignment we also asked how the output $2/6$ can be changed to
+the $3/6$. The outcome of this puzzle will be analyzed later in Section \ref{problem-sheets}.
+
+It was expected that the students solve the sheet problems on their own, without
+any communication. However, external help like using Internet was allowed for them.
+The students were informed that about 6--8 sheets should be solved in the project,
+and the required time for each should not exceed one hour.
 
 In fact, we created a couple of other programs in advance, but always waited for
-all solutions turned in by the students before finalizing the next set of problems.
+all solutions turned in by the students before finalizing the next sheet of problems.
 We \textit{did not} give any feedback on their solutions. Instead, if we felt that
-some difficulties arose, then we tried to cover them in the next set of problems.
+some difficulties arose, then we tried to cover them in the next sheet of problems.
 The reason behind this was to test if it is possible to create a set of problems
@@
 
-\section{The problem sets}
-\label{problem-sets}
+\section{The problem sheets}
+\label{problem-sheets}
+
+\subsection*{Sheet 1}
+The first sheet has already been mentioned in Section \ref{preps}.
+The expected explanation of the Python program was solved well by the students.
+All of them managed to write a correct answer to the creative question, namely
+\begin{lstlisting}[language=Python,firstnumber=5]
+  if Augenzahl > 3:
+\end{lstlisting}
+One student mentioned that the same output can be obtained with this change:
+\begin{lstlisting}[language=Python,firstnumber=2]
+günstig = 1
+\end{lstlisting}
+when line 5 is kept in its original form. Of course, this answer makes sense
+only from the programmer's point of view, and it has less to do with probability.
+
+After receiving the results for the first sheet we had the impression that
+the students got familiar with the basics of the required technology.
+
+\subsection*{Sheet 2}
+
+The second sheet consisted of the following four questions:
+
+\begin{enumerate}
+\item Consider the following program:
+\lstinputlisting[language=Python]{../progs/en_US/2.py}
+
+\item Explain its meaning line by line, in one sentence for each.
+
+\item Give the probability space where the possible outcomes are counted.
+Which event has been identified here as ``desired''?
+
+\item Modify the program to get a mechanical solution for the following question:
+
+\begin{quote}
+One fair die is rolled three times. Compute the probability for the event that the outcomes
+correspond to a monotone increasing sequence.
+\end{quote}
+
+\end{enumerate}
+
+We asked the students to try to solve the sheet problems without using any external help.
+In case of lack of success
+they were allowed to use a website that explains how to use if-then-else conditions in Python.
+
+All students faced the quoted question during the introductory course in probability, but that time the solution
+was expected to follow a purely pencil-and-paper method.
+
+\subsubsection*{The results}
+
+This sheet was much more difficult than the first one. One student reported that about an hour was required to solve it.
+We learned that one student has a misconception by mixing relative frequency and probability---it was
+a kind of surprise for this student that the program computes \textit{probability}, even if
+there is no randomness built in. (In this case the students' misconception was corrected
+by an explanation given in an email answer. Here we made an exception and forced
+communication to clarify this basic concept.)
+
+In general we were satisfied with the students' answers. All of them learned how three loops can
+be nested to iterate on $\Omega=[1,2,\ldots,6]^3$. Multiple ways were found to express monotonicity,
+including both the construct with an \texttt{and} clause and the short form
+\texttt{outcome_die_1 <= outcome_die_2 <= outcome_die_3} which is one of Python's strengths.
+
+In our experience it turned out that with little work a definitely large probability space
+can be studied. In fact, manual check of $[1,2,\ldots,6]^3$ (or, equivalently, deriving the sum
+$1+3+6+10+15+21$) is much more time consuming than writing a suitable program.
 

### 2021-02-18 progs/en_US/2.py

@@ -1,0 +1,9 @@
+total = 0
+desired = 0
+for outcome_die_1 in [1,2,3,4,5,6]:
+  for outcome_die_2 in [1,2,3,4,5,6]:
+    total = total + 1
+    if outcome_die_1 + outcome_die_2 == 10:
+      desired = desired + 1
+print(f"{desired} of {total}")
+print(f"p = {desired/total}")

### 2021-02-18 paper/article.tex
Fix LaTeX compilation error
Add more sheets

@@
   {€}{{\euro}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
-  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1,
+  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1
+  {Ω}{{$\Omega$}}1
+  {ω}{{$\omega$}}1
+  {→}{{$\to$}}1
+  {ℝ}{{$\mathbb{R}$}}1,
   columns=fullflexible,
@@
 In general we were satisfied with the students' answers. All of them learned how three loops can
-be nested to iterate on $\Omega=[1,2,\ldots,6]^3$. Multiple ways were found to express monotonicity,
+be nested to iterate on $\Omega=\{1,2,\ldots,6\}^3$. Multiple ways were found to express monotonicity,
 including both the construct with an \texttt{and} clause and the short form
-\texttt{outcome_die_1 <= outcome_die_2 <= outcome_die_3} which is one of Python's strengths.
+\texttt{outcome\_die\_1 <= outcome\_die\_2 <= outcome\_die\_3} which is one of Python's strengths.
 
-In our experience it turned out that with little work a definitely large probability space
-can be studied. In fact, manual check of $[1,2,\ldots,6]^3$ (or, equivalently, deriving the sum
+In our experience it turned out that with little work a definitely large sample space
+can be studied. In fact, manual check of $\{1,2,\ldots,6\}^3$ (or, equivalently, deriving the sum
 $1+3+6+10+15+21$) is much more time consuming than writing a suitable program.
+
+\subsection*{Sheet 3}
+
+Fig.~\ref{3.py} shows the main code that was announced in the third sheet.
+
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/3.py}
+\caption{The third program (translated into English)}
+\label{3.py}
+\end{figure}
+
+With this program we introduced some handy techniques via \texttt{itertools.product}
+to define Cartesian powers. We pointed to
+the second sheet and asked the students to rewrite their programs by using the new techniques.
+We also asked them to display all desired outcomes on the screen (with the \texttt{print} command)
+in a sorted fashion (via the \texttt{list} datatype).
+
+We had positive feedback---they managed to understand the point. A student successfully generalized the 
+solution for the puzzle of the second sheet for 5 dice, even if it was not directly asked.
+
+\subsection*{Sheet 4}
+
+The fourth sheet consisted of the following four questions:
+
+\begin{enumerate}
+\item Consider the following program:
+\lstinputlisting[language=Python]{../progs/en_US/4.py}
+
+\begin{enumerate}
+\item Check if the program works properly. Save a screenshot for an evidence.
+\item Explain the difference between this program and the one in sheet 3.
+\end{enumerate}
+
+\item Modify the program on line 7 and 14 in order to solve this problem:
+\begin{quote}
+Alice und Bob roll a die four times consecutively.
+If the sum of the outcomes is between 7 and 14, Alice wins, otherwise Bob.
+For which player is the game advantageous?
+\end{quote}
+
+\item Change the rules of the game above to get a fair play.
+Verify the changed rules by writing a suitable program.
+
+\item Assume that lines 12--15 are replaced by the following ones:
+\begin{lstlisting}[language=Python,numbers=none]
+E = [ω for ω in Ω if X(ω) == 10]
+\end{lstlisting}
+Check that the same result can be obtained as before.
+Also, explain this short form by using a mathematical formula like a definition for a set.
+\end{enumerate}
+
+In this sheet we focused on the concept of random variables. To highlight their importance
+we defined a Python function \texttt{X($\omega$)}.
+
+\subsubsection*{The results}
+
+The students answered the questions generally quite well. Two of them remarked that
+the problem setting in question 2 was incomplete since line 10 has to be changed as well.
+
+It turned out that the ``copy-paste'' way of copying a program was technically not always
+straightforward---sometimes indentation was transferred improperly and some manual edit
+was required on the student's side.
+
+Fig.~\ref{42} shows a solution by a student for question 3. It is clear that this
+student successfully combined previous programming knowledge, theory and praxis. Similarly,
+two other students found the same idea to change the limits to have $7\leq X(\omega)\leq14$,
+and one of them tried several attempts to find better limits---without success.
+(Note that for an even number of dice there is no trivial solution by halving the
+sample space symmetrically in the middle. So, in some sense, question 3 is adequate to motivate
+experimenting.)
+
+\begin{figure}
+\begin{center}\includegraphics[width=1.0\textwidth]{42}
+\caption{A student's solution for question 3 in sheet 4}
+\label{42}
+\end{center}
+\end{figure}
+
+The last question was solved perfectly by 3 students, giving the answer
+$$E=\{\omega\;\mid\;\omega\in\Omega \land X(\omega)=10\}$$
+where $X(\omega)=\omega_1+\omega_2$. Here we concluded that the technical
+fact that Python's indexing starts from $0$, can be confusing for beginners,
+since in mathematics the usual indexing starts from $1$.
 

### 2021-02-18 progs/en_US/3.py

@@ -1,0 +1,20 @@
+import itertools
+# 1. The basic set:
+outcomes = {1, 2, 3, 4, 5, 6}
+# itertools.product creates a Cartesian product:
+Ω = set(itertools.product(outcomes, outcomes))
+print(Ω)
+# Note that Ω as a set and it is unsorted.
+# Instead of "set" we can use a "list":
+# in that case the data will be sorted.
+
+# 2. Count the amount of desired and total outcomes:
+desired = 0
+total = len(Ω)
+for outcome_die_1, outcome_die_2 in Ω:
+  if outcome_die_1 + outcome_die_2 == 10:
+    desired = desired + 1
+
+# 3. Output:
+print(f"{desired} of {total}")
+print(f"p = {desired/total}")

### 2021-02-18 progs/en_US/4.py

@@ -1,0 +1,18 @@
+import itertools
+
+def X(ω):
+    """Students could edit this function on their own.
+    X is a rendom variable, that is, X: Ω → ℝ.
+    A real number will be mapped to each outcome."""
+    return ω[0]+ω[1]
+
+outcomes = {1, 2, 3, 4, 5, 6}
+Ω = list(itertools.product(outcomes, outcomes))
+# Event E
+E = []
+for ω in Ω:
+    if X(ω) == 10:
+        E.append(ω)
+print(f"{len(Ω)} total outcomes")
+print(f"{len(E)} desired outcomes")
+print(f"p = {len(E)/len(Ω)}")

### 2021-02-19 paper/article.tex
Add additional text, figures and program translations

@@
 
+\subsection*{Sheet 5}
+
+Fig.~\ref{5a.py} shows the first program in the fifth sheet. The students' assignment was
+to find a text problem that is solved by the program.
+
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/5a.py}
+\caption{The first program in the fifth sheet (translated into English)}
+\label{5a.py}
+\end{figure}
+
+To find a connection with binomial coefficients, the students had to explain why the result
+equals to $\binom{5}{2}/2^5$. A technical question helped in improving the look of the output,
+by suggesting the addition of this line between lines 2 and 3:
+\begin{lstlisting}[language=Python,numbers=none]
+Ω = {"".join(outcome) for outcome in Ω}
+\end{lstlisting}
+
+Then a second program was introduced. In fact, it was just a slight modification of the first program,
+but, for simplicity, the whole code was announced (see Fig.~\ref{5b.py}).
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/5b.py}
+\caption{The second program in the fifth sheet (translated into English)}
+\label{5a.py}
+\end{figure}
+Here the assignment was to give a random variable $X$ whose expected value is computed in the program.
+Another assignment was to explain why the expected value $2.5$ is printed. Then, this result
+had to be generalized by changing the code \texttt{repeat=5} to other values. Finally,
+also the variance had to be computed by adding some more lines to the code.
+
+\subsubsection*{The results}
+
+The first program was well identified by all students---a fair coin was tossed 5 times,
+and the probability of the appearance of exactly two heads was computed.
+Applying the technical hint was unsuccessful for two students (for one the closing brace was missing,
+for another one the issue was unclear).
+
+For the second program, one student added the line \texttt{print(len($\Omega$))} to better understand the
+code. (In fact, the function \texttt{len} was already introduced in the third sheet.) Instead
+of mentioning symmetry, the same student argued by citing the formula $E=n\cdot p$ for the
+binomial distribution where $n$ denotes the number of independent experiments and $p$ the probability
+of success (here $0.5$). The same student computed the variance by multiplying $E$ with $0.5$---this
+is correct for the given assignment, but here no programming was used, just theory.
+
+The other students explained the result $2.5$ by mentioning symmetry, and the same argument was
+used for the generalization. But no answer was found to compute the variance, only for one student,
+and in that case it was incorrect (see Fig.\ref{52})---clearly, on line 10 something like
+\begin{lstlisting}[language=Python,firstnumber=10]
+  Varianz = Varianz + (w.count("K")-Erwartungswert)**2
+\end{lstlisting}
+should have been written. Actually, the notation \texttt{**2} could be something unfamiliar
+for those who never did any programming before. Anyway, the result $0.078125$ should have been
+suspicious. It is also remarkable that this student preferred the Latin alphabet to Greek characters.
+\begin{figure}
+\begin{center}\includegraphics[width=1.0\textwidth]{52}
+\caption{A student's incorrect program code to compute variance in sheet 5}
+\label{52}
+\end{center}
+\end{figure}
+
+\subsection*{Sheet 6}
+
+We identified the assignment to compute the variance in the fifth sheet somewhat unsuccessful.
+So we continued by sharing two correct solutions to compute the variance (see Fig.~\ref{6a.py}
+and \ref{6b.py}).
+
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/6a.py}
+\caption{A correct solution to compute the variance (translated into English)}
+\label{6a.py}
+\end{figure}
+
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/6b.py}
+\caption{Another correct solution to compute the variance (translated into English)}
+\label{6b.py}
+\end{figure}
+
+The students had to compare these two program codes, check if their outputs are the same
+(even for 6 or 7 tosses), and find the differences between the way of programming.
+We actually wanted to add another connection with theory, namely, to verify if the
+formula
+\begin{equation}
+V(X)=E((X-E(X))^2)=E(X^2)-E(X)^2
+\label{variance}
+\end{equation}
+can be reflected in the two solutions.
+
+As another activity, a third program was given (see Fig.~\ref{6c.py}) to prepare the last
+topic, namely, some steps towards the central limit theorem (CLT).
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/6c.py}
+\caption{A third program in the sixth sheet (translated into English)}
+\label{6c.py}
+\end{figure}
+Here we provide the questions to this third program:
+\begin{enumerate}
+\item What computation is performed by this program?
+\item Consider following modifications:
+\begin{lstlisting}[language=Python,firstnumber=2]
+Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=4))
+possible_values = range(30)
+\end{lstlisting}
+\begin{lstlisting}[language=Python,firstnumber=8]
+  heads[sum(ω)] += 1
+\end{lstlisting}
+Explain the output of the modified program.
+\item We are about to plot the obtained data. So we change line 10 as follows:
+\begin{lstlisting}[language=Python,firstnumber=10]
+for w in zip(*(possible_values, probabilities)):
+  print(f"{w[0]},{w[1]}")
+\end{lstlisting}
+Copy the output of this program in a spreadsheet software and create an XY diagram
+to get some graphical output.
+\item Extend the program with these two additional lines:
+\begin{lstlisting}[language=Python,firstnumber=12]
+distribution = [sum(probabilities[:h]) for h in possible_values]
+print(list(zip(*(possible_values, distribution))))
+\end{lstlisting}
+Explain the result.
+\end{enumerate}
+In fact, the variable naming was somewhat incorrect here (because \texttt{heads} 
+actually refers to the sums of the outcomes), but we wanted to make very little
+changes on the code.
+
+Among the authors there was a quite a long discussion on getting the output in
+the best fashion. The Python construction \texttt{list(zip(*(...)))} is a kind
+of compromise between beautiful output with much code and simple output with less code,
+however, this is something non-trivial in Python.
+Finally a graphical output was also expected---students should be familiar with
+popular spreadsheet software and visualize data in an easier way than programming.
+
+\subsubsection*{The results}
+
+For the first part, all solutions were correct: three of them mentioned that
+for the second code only one loop is required, so it is somewhat more effective.
+In fact all of the turned-in solutions lacked of the wow feeling that we used
+the formula (\ref{variance}). One student reported
+experiments to learn if \texttt{**} is a typo or not---finally the correct interpretation
+was found: it is \textit{the power sign}.
+
+For the second part, the students successfully identified the purpose of the third program
+(that is, to find all probabilities for the different number of heads for 5 tosses), and of the changed one
+(to find all probabilities of the sums of the outcomes for 4 dice). Both the density
+function and the cumulative distribution were obtained correctly, and for the former,
+a graphical output was always attached. (One student created a graph for the latter as well.)
+We also learned that, in the case of three students, the density function of the
+\textit{discrete} binomial distribution
+was incorrectly identified, namely, as a \textit{continuous} normal distribution.
+
+We also noted that the modification \texttt{possible\_values = range(30)} aroused interest
+among the students,
+namely, to find the meaning of the number 30. (In fact, only the sums $4,5,\ldots,24$
+had a meaning, but for simplicity we wanted to avoid any manipulation of the indexes of
+the arrays.)
+
+\subsection*{Sheet 7}
+
+In the final sheet we wanted to point towards CLT, and thus we gave two programs:
+they can be seen in Fig.~\ref{7a.py} and \ref{7b.py}).
+
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/7a.py}
+\caption{First program in the seventh sheet (translated into English)}
+\label{7a.py}
+\end{figure}
+
+\begin{figure}
+\lstinputlisting[language=Python]{../progs/en_US/7b.py}
+\caption{Second program in the seventh sheet (translated into English)}
+\label{7b.py}
+\end{figure}
+
+In fact, these two programs are almost the same, just the outputs differ: the second
+one results in a nice diagram via \texttt{matplotlib} (see Fig.~\ref{72a}). We provided an alternative
+way to plot a vertical bar chart with the following suggestion:
+\begin{lstlisting}[language=Python,firstnumber=10]
+  p.bar(range(1, max_tosses * 6 + 1), probabilities[1:], alpha=0.5)
+\end{lstlisting}
+This is mathematically more correct because the discrete values are highlighted
+(see Fig.~\ref{72b}).
+On the other hand, to point towards CLT, also a continuous diagram makes sense.
+
+\begin{figure}
+\centering
+\subfloat[An XY diagram via \texttt{pyplot.plot}]{%
+\resizebox*{6cm}{!}{\includegraphics{72a}}\label{72a}}\hspace{5pt}
+\subfloat[A bar chart via \texttt{pyplot.bar}]{%
+\resizebox*{6cm}{!}{\includegraphics{72b}}}
+\caption{Graphical outputs of the second program in the seventh sheet} \label{sample-figure}
+\end{figure}
+
+Technically, we asked the students to use
+Sage worksheets at \url{cocalc.com} since it provides a simple interface to obtain
+the graphical result.
+
+In this closing activity we requested to obtain as many mathematical properties of the bar chart
+as possible. We asked the students to generalize their experiences. We wanted them to
+focus on using notions which are also familiar in a secondary school context. We clarified
+that notions like sequence, limit, function, density function, expected value, variance,
+standard deviation and normal distribution are all allowed.
+
+\subsubsection*{The results}
+
 \section*{Acknowledgement(s)}

### 2021-02-19 progs/en_US/4.py

@@ -1,18 +1,18 @@
 import itertools
 
 def X(ω):
-    """Students could edit this function on their own.
-    X is a rendom variable, that is, X: Ω → ℝ.
-    A real number will be mapped to each outcome."""
-    return ω[0]+ω[1]
+  """Students could edit this function on their own.
+  X is a rendom variable, that is, X: Ω → ℝ.
+  A real number will be mapped to each outcome."""
+  return ω[0]+ω[1]
 
 outcomes = {1, 2, 3, 4, 5, 6}
 Ω = list(itertools.product(outcomes, outcomes))
 # Event E
 E = []
 for ω in Ω:
-    if X(ω) == 10:
-        E.append(ω)
+  if X(ω) == 10:
+    E.append(ω)
 print(f"{len(Ω)} total outcomes")
 print(f"{len(E)} desired outcomes")
 print(f"p = {len(E)/len(Ω)}")

### 2021-02-19 progs/en_US/5a.py

@@ -1,0 +1,7 @@
+import itertools
+Ω = list(itertools.product({"h", "t"}, repeat=5))
+print(f"Ω={Ω}")
+E = [ω for ω in Ω if ω.count("h") == 2]
+print(f"E={E}")
+print(f"{len(E)} of {len(Ω)}")
+print(f"p = {len(E)/len(Ω)}")

### 2021-02-19 progs/en_US/5b.py

@@ -1,0 +1,7 @@
+import itertools
+Ω = set(itertools.product({"h", "t"}, repeat=5))
+expected_value = 0
+for ω in Ω:
+  expected_value = expected_value + ω.count("h")
+expected_value = expected_value / len(Ω)
+print(f"expected value = {expected_value}")

### 2021-02-19 progs/en_US/6a.py

@@ -1,0 +1,12 @@
+import itertools
+Ω = set(itertools.product({"h", "t"}, repeat=5))
+E = 0
+for ω in Ω:
+  E = E + ω.count("h")
+E = E / len(Ω)
+V = 0
+for ω in Ω:
+  V = V + (ω.count("h")-E)**2
+V = V / len(Ω)
+print(f"expected value = {E}")
+print(f"variance = {V}")

### 2021-02-19 progs/en_US/6b.py

@@ -1,0 +1,12 @@
+import itertools
+Ω = set(itertools.product({"h", "t"}, repeat=5))
+E = 0
+V = 0
+for ω in Ω:
+  x = ω.count("h")
+  E = E + x
+  V = V + x*x
+E = E / len(Ω)
+V = V / len(Ω) - E*E
+print(f"expected value = {E}")
+print(f"variance = {V}")

### 2021-02-19 progs/en_US/6c.py

@@ -1,0 +1,10 @@
+import itertools
+Ω = set(itertools.product({"h", "t"}, repeat=5))
+possible_values = [0, 1, 2, 3, 4, 5]
+# In the beginning each value of the random variable
+# has a frequency of 0.
+heads = [0]*len(possible_values)
+for ω in Ω:
+  heads[ω.count("h")] += 1
+probabilities = [h/len(Ω) for h in heads]
+print(list(zip(*(possible_values, probabilities))))

### 2021-02-19 progs/en_US/7a.py

@@ -1,0 +1,9 @@
+import itertools
+max_tosses = 4
+for W in range(1, max_tosses + 1):
+  Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
+  sums = [0] * (max_tosses * 6 + 1)
+  for ω in Ω:
+    sums[sum(ω)] += 1
+  probabilities = [h/len(Ω) for h in sums]
+  print(probabilities)

### 2021-02-19 progs/en_US/7b.py

@@ -1,0 +1,11 @@
+import itertools
+import matplotlib.pyplot as p
+max_tosses = 4
+for W in range(1, max_tosses + 1):
+  Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
+  sums = [0] * (max_tosses * 6 + 1)
+  for ω in Ω:
+    sums[sum(ω)] += 1
+  probabilities = [h / len(Ω) for h in sums]
+  p.plot(range(1, max_tosses * 6 + 1), probabilities[1:])
+p.show()

### 2021-02-19 paper/article.tex
Fix internal reference
Add the remaining sections

@@
 
-%\begin{figure}
-%\lstinputlisting[language=Python]{../progs/de_AT/1.py}
-%\caption{The first program in its original form (in German)}
-%\label{1.py}
-%\end{figure}
-
 The first example is shown in Fig.~\ref{main-py-window}. We kept the original version to highlight
@@
 \begin{figure}
-\begin{center}\includegraphics[scale=0.5]{../forms/de_AT/main-py-window}
+\begin{center}\includegraphics[scale=0.4]{../forms/de_AT/main-py-window}
 \caption{An explanatory screenshot to help the students' first steps}
@@
 The last question was solved perfectly by 3 students, giving the answer
-$$E=\{\omega\;\mid\;\omega\in\Omega \land X(\omega)=10\}$$
+$$E=\{\omega\;\mid\;\omega\in\Omega\;\land\;X(\omega)=10\}$$
 where $X(\omega)=\omega_1+\omega_2$. Here we concluded that the technical
@@
 \caption{The second program in the fifth sheet (translated into English)}
-\label{5a.py}
+\label{5b.py}
 \end{figure}
@@
 The students had to compare these two program codes, check if their outputs are the same
-(even for 6 or 7 tosses), and find the differences between the way of programming.
+(even for 6 or 7 dice), and find the differences between the way of programming.
 We actually wanted to add another connection with theory, namely, to verify if the
@@
 \end{equation}
+(also known as Steiner's \textit{Verschiebungssatz}, that is, \textit{translation theorem})
 can be reflected in the two solutions.
@@
 For the second part, the students successfully identified the purpose of the third program
-(that is, to find all probabilities for the different number of heads for 5 tosses), and of the changed one
+(that is, to find all probabilities for the different number of heads for 5 dice), and of the changed one
 (to find all probabilities of the sums of the outcomes for 4 dice). Both the density
-function and the cumulative distribution were obtained correctly, and for the former,
+function and the cumulative distribution function (CDF) were obtained correctly, and for the former,
 a graphical output was always attached. (One student created a graph for the latter as well.)
@@
 \begin{lstlisting}[language=Python,firstnumber=10]
-  p.bar(range(1, max_tosses * 6 + 1), probabilities[1:], alpha=0.5)
+  p.bar(range(1, max_dice * 6 + 1), probabilities[1:], alpha=0.5)
 \end{lstlisting}
@@
 \subfloat[A bar chart via \texttt{pyplot.bar}]{%
-\resizebox*{6cm}{!}{\includegraphics{72b}}}
+\resizebox*{6cm}{!}{\includegraphics{72b}}\label{72b}}
 \caption{Graphical outputs of the second program in the seventh sheet} \label{sample-figure}
@@
 
-\section*{Acknowledgement(s)}
+The students managed to visualize both diagrams and most of them concluded that the bar chart is
+more useful to study the exact values of the functions. A student, however, preferred the XY diagram
+that may provide more precise values.
+They managed to read off the following relationships:
+\begin{enumerate}
+\item The definite integral of the density function is $1$ (between $0$ and $25$).
+\item The maximum of the density function is reached at the expected value. There is a clear
+relationship between the expected value and the number of rolls.
+\end{enumerate}
+
+According to CLT, the following statements were claimed (here and from now on we use $d$
+for the number of dice):
+\begin{itemize}
+\item ``The more rolls you perform, the clearer the normal distribution of the [density] function becomes.''
+\item For $d=3$ and $d=4$ ``it is clear that it is about the normal distribution of the sums,
+because of its beautiful bell shaped curve.''
+\item For $d=3$ ``we already obtain an approximate Gaussian bell curve (normal distribution).''
+\item For $d=4$ ``we can already see a beautiful normal distribution with the expected value 14.''
+\end{itemize}
+
+Unfortunately, the notion ``limit'' was mixed with the notion ``maximum'' in some sense, so two students did not
+observe at all how the situation changes if $d\to\infty$. One student studied each density function
+in the case $h\to\infty$ (cf.~line 9 in Fig.~\ref{7b.py}), to obtain its limit, which is, of course, $0$.
+(Also the same student dealt with the case $h\to-\infty$.) Finally, this student used 
+the $68.27\%$ rule (by using external information from a different course on
+probability theory) to get the standard deviation in the case $d=4$,
+under the assumption that $i$ is an inflection point of the density function:
+\begin{equation}
+\Phi\left(\frac{i-E(X)}{\sqrt{V(X)}}\right)=0.6827,
+\end{equation}
+where $\Phi$ denotes the CDF of the standard normal distribution, and the inflection point is guessed
+to be at $i=18$. With this approximation formula the value $\sqrt{V(X)}\approx8.33$ can be obtained.
+
+\section{Discussion}
+
+We did not have any preliminary goals at the beginning of this project, only to connect theory and
+practice via programming. After evaluating the solutions we have the feeling that all students
+have a safer approach to problem solving in probability theory, and they have no fear to give
+a computer program a chance to do experiments. That is, even if they need to modify the program code,
+studying via programming experiments is a feasible way.
+
+Our project confirmed that Python is a very useful and practical tool to collect exact data
+to study probability theory---even if there can appear some minor inconveniences.
+As one student pointed out, by using Sage, symbolic computation
+supports even more precise observations---for example, the first program in the last sheet
+(Fig.~\ref{7a.py}) was accidentally launched in Sage as well and all results were shown
+as vulgar fractions.
+
+On the other hand, connecting theory and the result of a program was not always straightforward.
+Students identified the CDF very well, but they were unable to recognize the CLT, even if
+it was discussed during the former course---but maybe without a result of deeper understanding.
+On the other hand, the CLT was not directly asked in the seventh sheet,
+so the problem setting was maybe not clear enough.
+
+We conjecture that making a difference between a discrete distribution and a continuous one may be
+a big jump in the level of understanding for the student. Discrete distributions are \textit{never} normal,
+and to speak about a Gaussian distribution one should introduce the notion of limit in a very
+delicate form. Here, we should translate the random variable $X$ to have the expected value $0$ first,
+and then normalize it by a function of $d$. We did not see any sign of this concept among the
+students' solutions. Maybe, in a next project, we will explicitly ask the students to try to
+\textit{approximate} the discrete distributions with a continuous one, by searching for the adequate
+expected value and standard deviation. Also, experimenting with higher values for $d$ should also be
+suggested---no student
+tried to increase the value of \texttt{max\_dice}, even if technically it is possible to try
+\texttt{max\_dice = 8} in a \texttt{cocalc.com} Sage session. (Fig.~\ref{10} shows the output
+for \texttt{max\_dice = 10} when run natively on a recent computer, after about 1 minutes
+of computation. Here we warn the reader that a sample space $\Omega$ with more than 60 millions of elements
+must be observed, since $|\Omega|=6^{10}=60,466,176$.)
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.7\textwidth]{10}
+\caption{Increasing the value of \texttt{max\_dice} to $10$}
+\label{10}
+\end{center}
+\end{figure}
+
+Finally, we remark that experimenting with a non-fair die could be another step to explain
+the CLT in a more general approach. A fair die, after rolled three or four times,
+results in a ``too symmetrical'' binomial distribution---it can be confused with
+the normal distribution. A non-symmetrical result, which is still similar enough
+to the normal distribution to make a conjecture, could be eventually a better
+example to promote understanding the CLT and help students in formalizing it---just
+by using their experiments and own words.
+(For a nice continuous example we refer to the Wikipedia page
+\url{https://en.wikipedia.org/wiki/Illustration_of_the_central_limit_theorem#Illustration_of_the_continuous_case}.)
+
+\section{Conclusion}
+
+We reported on an experimental project that connects an email based Python course
+and former knowledge of elementary concepts of probability theory.
+We think our results successfully present that there is room for this kind of approach
+in the education. We also think that our method should be further refined and tested.
+
+\section*{Acknowledgements}
+
+We are thankful to our students Alexander, Romana, Sandra and Sarah
+for their participation in the project.
 

### 2021-02-19 progs/en_US/7a.py

@@ -1,8 +1,8 @@
 import itertools
-max_tosses = 4
-for W in range(1, max_tosses + 1):
-  Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
-  sums = [0] * (max_tosses * 6 + 1)
+max_dice = 4
+for d in range(1, max_dice + 1):
+  Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=d))
+  sums = [0] * (max_dice * 6 + 1)
   for ω in Ω:
     sums[sum(ω)] += 1
   probabilities = [h/len(Ω) for h in sums]

### 2021-02-19 progs/en_US/7b.py

@@ -1,11 +1,11 @@
 import itertools
 import matplotlib.pyplot as p
-max_tosses = 4
-for W in range(1, max_tosses + 1):
-  Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=W))
-  sums = [0] * (max_tosses * 6 + 1)
+max_dice = 4
+for d in range(1, max_dice + 1):
+  Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=d))
+  sums = [0] * (max_dice * 6 + 1)
   for ω in Ω:
     sums[sum(ω)] += 1
   probabilities = [h / len(Ω) for h in sums]
-  p.plot(range(1, max_tosses * 6 + 1), probabilities[1:])
+  p.plot(range(1, max_dice * 6 + 1), probabilities[1:])
 p.show()

### 2021-02-19 paper/article.tex
Fix typo

@@
 used for the generalization. But no answer was found to compute the variance, only for one student,
-and in that case it was incorrect (see Fig.\ref{52})---clearly, on line 10 something like
+and in that case it was incorrect (see Fig.~\ref{52})---clearly, on line 10 something like
 \begin{lstlisting}[language=Python,firstnumber=10]

### 2021-03-13 paper/article.tex
Update the paper according to Alex's suggestion, add references to literature

@@
 sense, statistics is much easier to explain---one needs to run virtual experiments
-in a web browser % add links to Steve Phelps' https://www.geogebra.org/m/bxytp4hq, Ice Cream
-% https://www.geogebra.org/m/qsjbqjfe, Broken Stick
-% https://www.geogebra.org/m/bu5sxbn2, Fair Coin
-% + Andreas Lindner's collections: https://www.geogebra.org/m/AytaSakt, https://www.geogebra.org/m/qXPnyCTZ
-and this is well supported for several years, among others, in the online version of GeoGebra.
+in a web browser
+and this is well supported for several years, among others, in the online version of GeoGebra\footnote{%
+See, for example, Steve Phelps' online applets
+\url{https://www.geogebra.org/m/bxytp4hq},
+\url{https://www.geogebra.org/m/qsjbqjfe} and
+\url{https://www.geogebra.org/m/bu5sxbn2},
+that visualize experiments with ice cream purchases, a broken stick or a fair coin, respectively;
+or check out Andreas Lindner's collections at \url{https://www.geogebra.org/m/AytaSakt}
+and \url{https://www.geogebra.org/m/qXPnyCTZ}.}.
 
@@
 theory and practice, but their students either. Roughly speaking: everybody talks about probability and
-statistics, but nobody understands anything behind the scenes.
+statistics, but nearly nobody understands anything behind the scenes.
 
-In our paper we propose a radical way, or at least, to extend the classic way how
+In our paper we propose a radical way, or at least an extension of the classic way, how
 basics of elementary probability theory should be taught. It is \textit{programming}.
 Our conjecture is that understanding will be significantly improved if the main concepts
-of probability theory are supported with simple computer programs. The ouput for their flexible input
+of probability theory are supported with simple computer programs. The output for their flexible input
 should be immediately checked, eventually in a web browser, and by changing the input parameters
@@
 First author hold a series of supplementary lectures on probability theory two years ago, where second author
-took part a student and illustrated several homework assignments by using Python programs.
+took part as a student and illustrated several homework assignments by using Python programs.
 Some of his illustrations were a bit complicated for beginners, so the idea came that
@@
 
-The first example is shown in Fig.~\ref{main-py-window}. We kept the original version to highlight
+\subsection{Why Python?}
+
+Why did we choose Python as the programming language for this course?
+There are a couple of reasons we would like to note. First of all: It is
+free, independent of the used operating system, has good documentation
+and great tutorials. In addition, Python is currently a very popular
+programming language: according to
+\url{https://madnight.github.io/githut/#/pull_requests/2020/4}, Python
+was the second most popular language on GitHub in the 4th quarter of 2020;
+moreover, ZDNet's recent article \cite{LiamTung2020} highlights that Python
+was the second best popular language in November 2020.
+Many programming courses teach it as a first
+programming language because of its ease of use and short expressive
+code.
+
+Also, we think that Python is a great tool to express mathematical
+concepts in a concise and direct way---several recent books give
+examples on a possible start, see e.g.~\cite{Fuhrer2016}, \cite{Mehta2015} and \cite{Stewart2014}.
+In fact, in our course students
+needed to learn classic programming concepts like conditions and loops
+together with probability, so we needed a programming language with a
+steep learning curve. We think Python can be seen as a great tool that
+supports students with the understanding of mathematical probability
+concepts without the need to learn too much programming.
+
+On the other hand, in the field of data science
+Python is currently the mostly used programming language. Especially in
+machine learning, which could be seen as an advanced application of
+probability, Python is very popular. We think that students who believe
+that the exercises they solve are not artificially constructed for the
+course, but have some connection to the real world, will be more
+motivated. As a teacher we can give students the impression that if they
+use Python in their introductory course on probability, they take a first
+step into the prestigious field of data science. This may increase the
+motivation of students and as a result they will achieve higher.
+
+\subsection{The programs}
+
+The first example of the introduced Python programs
+is shown in Fig.~\ref{main-py-window}. We kept the original version to highlight
 the possibility to write special characters in the Python code (we used Python version 3),
@@
 We also instructed all students with an
-explanatory screenshot how the program should be started and the output obtained.
+explanatory screenshot how the program should be started and the output can be obtained.
 
@@
 In fact, we created a couple of other programs in advance, but always waited for
-all solutions turned in by the students before finalizing the next sheet of problems.
+all solutions to be turned in by the students before finalizing the next sheet of problems.
 We \textit{did not} give any feedback on their solutions. Instead, if we felt that
@@
 played a significant role in our project. We did not want to ask the students
-to install anything on their computer. Instead, we tried to minimize the gap
+to install anything on their computers. Instead, we tried to minimize the gap
 between theory and practice. In fact, \url{repl.it} is not the only website
@@
 \end{equation}
-(also known as Steiner's \textit{Verschiebungssatz}, that is, \textit{translation theorem})
+(also known as Steiner translation theorem in Germany and König-Huygens formula in the French literature)
 can be reflected in the two solutions.
@@
 As another activity, a third program was given (see Fig.~\ref{6c.py}) to prepare the last
-topic, namely, some steps towards the central limit theorem (CLT).
+topic, namely, some steps towards the central limit theorem (CLT, \cite{Polya1920}).
 \begin{figure}
@@
 
-We conjecture that making a difference between a discrete distribution and a continuous one may be
+At this point we need to give some description on which form of CLT should be expected to discover by the students.
+In fact, CLT has several versions: both formal and informal variants. Roughly speaking, CLT can be summarized
+in the following form (see \cite{Wikipedia-CLT}):
+\begin{quote}
+When independent random variables are added, their properly normalized
+sum tends toward a normal distribution (informally a bell curve) even if
+the original variables themselves are not normally distributed.
+\end{quote}
+A formal definition given in \cite[p.~241]{Montgomery2014} is as follows:
+\begin{quote}
+If $X_{1},X_{2},\ldots,X_{n}$
+are $n$ random samples drawn from a population
+with overall mean $\mu$ and finite variance $\sigma^2$,
+and if ${\bar {X}}_{n}$ is the sample mean, the
+limiting form of the distribution, $$Z=\lim _{n\to \infty}{\sqrt {n}}{\left({\frac {{\bar {X}}_{n}-\mu }
+{\sigma}}\right)},$$ is the standard normal
+distribution.
+\end{quote}
+In fact, we avoided going into any formal detail during our course by assuming that the provided formula
+here \textit{may not promote understanding}.
+
+More importantly, we conjecture that making a difference between a discrete distribution and a continuous one may be
 a big jump in the level of understanding for the student. Discrete distributions are \textit{never} normal,
@@
 
+We emphasize that this research just the first step in a longer research project.
+Basically, we would like to continue with interviews with the participating
+students whether their knowledge and motivation were improved during the course.
+This will give us further feedback to meet the best decision towards further
+refinements of the course.
+
 \section*{Acknowledgements}
@@
 
-\begin{thebibliography}{99}
-
-\end{thebibliography}
+\bibliographystyle{tfs}
+\bibliography{article}
 

### 2021-03-13 paper/article.tex
Fix minor details
Fix typo

@@
 \texttt{max\_dice = 8} in a \texttt{cocalc.com} Sage session. (Fig.~\ref{10} shows the output
-for \texttt{max\_dice = 10} when run natively on a recent computer, after about 1 minutes
+for \texttt{max\_dice = 10} when run natively on a recent computer, after about 1 minute
 of computation. Here we warn the reader that a sample space $\Omega$ with more than 60 millions of elements
@@
 
-We are thankful to our students Alexander, Romana, Sandra and Sarah
+We are thankful to the students Alexander, Romana, Sandra and Sarah
 for their participation in the project.

### 2021-04-06 paper/article.tex
Add link to online materials
Update text after Andreas Lindner's suggestions

@@
 thus it is also an important item in the teacher training. But, to our experience,
-most prospective teachers \textit{never} understand the basics beyond introductory level.
+many prospective teachers have difficulties \textit{even with the basics} beyond introductory level.
 Our conjecture is that the main problem lies in students' verification if theory indeed meets practice,
@@
 theory and practice, but their students either. Roughly speaking: everybody talks about probability and
-statistics, but nearly nobody understands anything behind the scenes.
+statistics, but just a few have \textit{real understanding} behind the scenes.
 
@@
 step into the prestigious field of data science. This may increase the
-motivation of students and as a result they will achieve higher.
+motivation of students and as a result they will achieve a higher level.
 
@@
 elegance and speed. We learned that Python itself could be improved to
-have simpler constructions than the \texttt{for i in range(...)} loop for iteration.
-At the end of day we were still satisfied with all of the code and the students'
+provide a simpler way to construct iterations than the usual \texttt{for i in range(...)} loop.
+At the end of the day we were still satisfied with all of the code and the students'
 solutions confirm that our choices were acceptable.
@@
 
+All materials of our project---including the Python programs, the homework assignments and
+the turned-in homework---can be found online at \url{https://github.com/kovzol/probability-python}.
+
 \section*{Acknowledgements}
@@
 We are thankful to the students Alexander, Romana, Sandra and Sarah
-for their participation in the project.
+for their participation in the project, and to Andreas Lindner who
+gave us useful feedback on the first version of the manuscript.
 

### 2022-05-04 paper/article.tex
Some grammatical improvements

@@
 to a higher number, the results can be quickly generalized. We emphasize that computers
-can take on the high number of computatations from human---and this is what we exactly want.
+can take on the high number of computations from human---and this is what we exactly want.
 A quite simple \textit{sample space} can contain \textit{a lot} of elements, even millions or more, and it cannot
@@
 \url{https://madnight.github.io/githut/#/pull_requests/2020/4}, Python
-was the second most popular language on GitHub in the 4th quarter of 2020;
+was the second most favoured language on GitHub in the 4th quarter of 2020;
 moreover, ZDNet's recent article \cite{LiamTung2020} highlights that Python
-was the second best popular language in November 2020.
+was the second best liked language in November 2020.
 Many programming courses teach it as a first
@@
 machine learning, which could be seen as an advanced application of
-probability, Python is very popular. We think that students who believe
+probability, Python is very well liked. We think that students who believe
 that the exercises they solve are not artificially constructed for the
@@
 \begin{quote}
-Alice und Bob roll a die four times consecutively.
+Alice and Bob roll a die four times consecutively.
 If the sum of the outcomes is between 7 and 14, Alice wins, otherwise Bob.
@@
 Fig.~\ref{42} shows a solution by a student for question 3. It is clear that this
-student successfully combined previous programming knowledge, theory and praxis. Similarly,
+student successfully combined previous programming knowledge, theory and practise. Similarly,
 two other students found the same idea to change the limits to have $7\leq X(\omega)\leq14$,
@@
 We are thankful to the students Alexander, Romana, Sandra and Sarah
-for their participation in the project, and to Andreas Lindner who
-gave us useful feedback on the first version of the manuscript.
+for their participation in the project, and to Andreas Lindner
+and two anonymous reviewers who
+gave us useful feedback on the first versions of the manuscript.
 

### 2022-05-17 paper/article.tex
Simplify zip

@@
 \begin{lstlisting}[language=Python,firstnumber=10]
-for w in zip(*(possible_values, probabilities)):
+for w in zip(possible_values, probabilities):
   print(f"{w[0]},{w[1]}")
@@
 distribution = [sum(probabilities[:h]) for h in possible_values]
-print(list(zip(*(possible_values, distribution))))
+print(list(zip(possible_values, distribution)))
 \end{lstlisting}
@@
 Among the authors there was a quite a long discussion on getting the output in
-the best fashion. The Python construction \texttt{list(zip(*(...)))} is a kind
+the best fashion. The Python construction \texttt{list(zip(...))} is a kind
 of compromise between beautiful output with much code and simple output with less code,

### 2022-05-17 progs/en_US/6c.py

@@ -7,4 +7,4 @@ heads = [0]*len(possible_values)
 for ω in Ω:
   heads[ω.count("h")] += 1
 probabilities = [h/len(Ω) for h in heads]
-print(list(zip(*(possible_values, probabilities))))
+print(list(zip(possible_values, probabilities)))
